---
title: "ppod: an overview"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ppod: an overview}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width = 6,
  fig.height = 4
)
set.seed(20260515)
```

## Introduction

The `ppod` package implements a Bayesian phase I/II adaptive design
for optimal biological dose (OBD) selection with continuous toxicity
and efficacy endpoints. The methodological contribution is a
**posterior predictive expected utility** criterion that integrates
parameter uncertainty, residual outcome variability, and within-dose
toxicity-efficacy dependence. The package implements three variants:

- `eu_joint`: posterior predictive utility under the joint Gaussian
  working model.
- `eu_marginal`: the same predictive utility evaluated with
  \(\rho_j \equiv 0\) at the utility integration step.
- `eu_plugin`: utility evaluated at posterior-mean dose-response
  estimates, ignoring residual variability and parameter uncertainty.

The contrast between joint and marginal isolates the role of
dependence-sensitive utility evaluation; the contrast between joint
and plug-in isolates the role of posterior predictive integration.

This vignette walks through the design on one scenario from the
manuscript.

```{r}
library(ppod)
```

## The utility function

The package uses a substitution utility on the bounded unit square,
\[
U(x_T, x_E) = x_E^{\alpha}\,\exp\!\bigl(-\lambda x_T^{\beta}\bigr),
\]
where \(x_T = g_T(Y_T)\) and \(x_E = g_E(Y_E)\) are toxicity and
efficacy on the bounded scale. Under \(\alpha, \beta \ge 1\) and
\(\lambda > 0\), the mixed partial \(U_{TE}\) is strictly negative on
\((0,1)^2\), which is the substitutability condition under which
Corollary 1 of the manuscript holds: expected utility strictly
decreases in within-dose toxicity-efficacy correlation.

```{r}
util <- make_utility(alpha = 1, beta = 1, lam = 2)
# Utility at (low tox, high efficacy) is close to 1
util(x_T = 0.05, x_E = 0.95)
# Utility at (high tox, high efficacy) is close to exp(-lambda)
util(x_T = 0.95, x_E = 0.95)
```

## A scenario

The 12 manuscript scenarios are available as `ppod_scenarios`. We use
scenario S4b: unimodal efficacy with mild downturn at \(d_5\),
constant within-dose correlation \(\rho = 0\), residual scale
\(\sigma = 0.5\).

```{r}
scen <- ppod_scenarios$S4b
data.frame(
  dose  = 1:5,
  mu_T  = scen$mu_T,
  mu_E  = scen$mu_E,
  x_T   = plogis(scen$mu_T),
  x_E   = plogis(scen$mu_E)
)
```

The true OBD under the data-generating distribution is computed by
Monte Carlo over the joint bivariate normal:

```{r}
eu_true <- true_expected_utility(scen)
data.frame(dose = 1:5, Psi_true = round(eu_true, 4))
which.max(eu_true)
```

The true OBD is \(d_3\), matching `scen$true_OBD`.

## One adaptive trial

`simulate_trial` runs one replicate of the adaptive design. The
default settings reproduce the manuscript: \(N = 54\) patients,
cohort size 3, run-in to \(n_0 = 9\), then utility-based assignment
on the predictive admissible set with no-skipping-upward.

```{r trial}
result <- simulate_trial(scen, criterion = "joint")
result$selected            # selected OBD index
table(factor(result$dose_history, levels = 1:5))
```

If the trial stopped without a recommendation, `result$selected` is
`NA_integer_` and one of `result$stopped_tox` or
`result$stopped_fut` is `TRUE`.

## Comparing the three continuous-endpoint criteria

The three continuous criteria (joint, marginal, plug-in) can be
compared on the same scenario by varying the `criterion` argument.
For speed in this vignette we use a small replicate count.

```{r compare, cache = FALSE}
designs <- c("joint", "marginal", "plugin")
cells <- lapply(designs, function(d) {
  run_replicates(scen, design = d, n_reps = 10L, seed_base = 1000L,
                  S_interim = 50L, M_interim = 20L,
                  S_final = 100L, M_final = 30L)
})
names(cells) <- designs

sapply(cells, function(c) {
  oc <- operating_characteristics(c, scen)
  c(p_correct = round(oc$p_correct, 2),
    mean_loss = round(oc$mean_loss, 4))
})
```

In production, replicate counts of 200 to 400 are needed to bring the
Monte Carlo standard error on `p_correct` below 3 percentage points.

## Comparing against BOIN12

The package also includes a simplified BOIN12 binary-endpoint
comparator (Lin et al. 2020). The induced binary toxicity and
efficacy probabilities are derived by thresholding the continuous
outcomes at the same clinically meaningful cutoffs used by the
proposed design's predictive admissibility gate.

```{r boin12, cache = FALSE}
cell_boin <- run_replicates(scen, design = "boin12", n_reps = 10L,
                              seed_base = 2000L)
oc_boin <- operating_characteristics(cell_boin, scen)
c(p_correct = round(oc_boin$p_correct, 2),
  mean_loss = round(oc_boin$mean_loss, 4),
  overdose  = round(oc_boin$alloc["over"], 3))
```

## The S4d stress test

The dose-varying dependence stress test S4d is the scenario where
Proposition 3 of the manuscript predicts a marginal-versus-joint OBD
reversal at the deterministic truth.

```{r s4d}
s4d <- ppod_scenarios$S4d
# Joint and marginal true OBDs differ
eu_j <- true_expected_utility(s4d)
eu_m <- true_expected_utility(s4d, marginal = TRUE)
data.frame(dose = 1:5,
           Psi_joint    = round(eu_j, 4),
           Psi_marginal = round(eu_m, 4))
which.max(eu_j); which.max(eu_m)
```

Joint OBD is \(d_2\); marginal OBD is \(d_3\). At \(N = 54\) the
adaptive trial typically cannot resolve this reversal reliably, since
the deterministic gap (about 0.02 on a utility range of about 0.27)
is small relative to within-trial posterior uncertainty; see Section 5
of the manuscript and the supplementary `inst/scripts/reproduce_supp.R`.

## Posterior inference at one cohort

For users who want to inspect the inferential machinery directly:

```{r posterior}
set.seed(1)
n_per <- 6L
doses <- rep(1:5, each = n_per)
J <- 5L
# Simulate a 30-patient dataset from S4b's data-generating model
Y_T <- stats::rnorm(J * n_per, mean = scen$mu_T[doses], sd = scen$sigma_T)
Y_E <- stats::rnorm(J * n_per, mean = scen$mu_E[doses], sd = scen$sigma_E)

fit <- fit_laplace(Y_T, Y_E, doses, J = J)
fit$ok

draws <- sample_posterior(fit$theta, fit$cov, S = 500)
util  <- make_utility(1, 1, 2)
data.frame(
  dose     = 1:J,
  eu_joint    = round(eu_joint(draws, M = 50, util, J = J), 4),
  eu_marginal = round(eu_marginal(draws, M = 50, util, J = J), 4),
  eu_plugin   = round(eu_plugin(fit$theta, util, J = J), 4)
)
```

## Where to go next

- The 12 simulation scenarios: `?ppod_scenarios`.
- Constructing a custom scenario: `?make_scenario`.
- Custom utility families: `?make_utility`.
- Full reproducibility scripts:
  `system.file("scripts", package = "ppod")`.
- Manuscript and supplementary material: see the project README.
